---
title: "Predation mortality analysis"
author: "Caliyena Brown"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
params:
  temperature_threshold: 20
  excluded_predation_fish: 72
---

```{r setup, include=FALSE}

library(dplyr)
library(tidyr)
library(readr)
library(lubridate)
library(ggplot2)
library(patchwork)
library(survival)
library(coxphf)
```

# Purpose

This document identifies temperature-defined predation events, calculates fate
and predation proportions, compares total length and sex composition between
release cohorts, and evaluates the effects of total length and sex on time to
inferred predation using a Firth-penalized Cox proportional hazards model.

# Import and validate data

```{r import-data}
det <- readRDS("det_clean_no_release_site_det.rds")
meta_tag <- readRDS("meta_clean.rds")

# Convert tag-level metadata to one record per fish
meta <- meta_tag %>%
  select(fish, sex, tl, fl, release_dt) %>%
  distinct() %>%
  mutate(
    cohort = case_when(
      lubridate::month(release_dt) == 4 ~ "April",
      lubridate::month(release_dt) == 12 ~ "December"
    ),
    cohort = factor(cohort, levels = c("April", "December")),
    sex = factor(
      sex,
      levels = c("F", "M"),
      labels = c("Female", "Male")
    )
  )

stopifnot(
  nrow(meta) == 101,
  anyDuplicated(meta$fish) == 0
)

det <- det %>%
  transmute(
    fish = as.integer(fish),
    temp = as.numeric(temp),
    dt = ymd_hms(dt, tz = "UTC", quiet = TRUE)
  ) %>%
  filter(!is.na(fish), !is.na(dt), fish %in% meta$fish) %>%
  arrange(fish, dt)
```

# Detection and fate summaries

```{r identify-events}
detected_fish <- intersect(meta$fish, unique(det$fish))
undetected_fish <- setdiff(meta$fish, detected_fish)

# The first record above 20 degrees C is treated as the inferred predation
# event. Fish 72 is excluded because review of detections during the same period
# showed that its high-temperature record was not a rapid, prolonged increase.
predation_events <- det %>%
  filter(
    !is.na(temp),
    temp > params$temperature_threshold,
    !fish %in% params$excluded_predation_fish
  ) %>%
  group_by(fish) %>%
  slice_min(dt, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  rename(predation_dt = dt, predation_temp = temp)

last_detections <- det %>%
  group_by(fish) %>%
  summarise(last_detection_dt = max(dt), .groups = "drop")

fish_fates <- meta %>%
  left_join(last_detections, by = "fish") %>%
  left_join(
    predation_events %>% select(fish, predation_dt, predation_temp),
    by = "fish"
  ) %>%
  mutate(
    detected = fish %in% detected_fish,
    predation = !is.na(predation_dt),
    fate = case_when(
      predation ~ "Inferred predation",
      !detected ~ "Mortality assumed: no post-release detection",
      TRUE ~ "Unknown"
    )
  )

fate_summary <- fish_fates %>%
  count(cohort, fate, name = "n") %>%
  group_by(cohort) %>%
  mutate(proportion_released = n / sum(n)) %>%
  ungroup()

predation_summary <- fish_fates %>%
  group_by(cohort) %>%
  summarise(
    n_released = n(),
    n_detected = sum(detected),
    n_predated = sum(predation),
    proportion_of_released = n_predated / n_released,
    proportion_of_detected = n_predated / n_detected,
    .groups = "drop"
  ) %>%
  bind_rows(
    fish_fates %>%
      summarise(
        cohort = factor("All", levels = c("April", "December", "All")),
        n_released = n(),
        n_detected = sum(detected),
        n_predated = sum(predation),
        proportion_of_released = n_predated / n_released,
        proportion_of_detected = n_predated / n_detected
      )
  )

knitr::kable(fate_summary, digits = 3,
             caption = "Fate classifications by release cohort.")
knitr::kable(predation_summary, digits = 3,
             caption = "Inferred predation as a proportion of released and detected kelts.")
```

# Detection histories and temperature profiles

Detection histories were visualized separately for the April and December
release cohorts using abacus plots. Points represent individual detections and
are coloured according to recorded transmitter temperature. For kelts with an
inferred predation event, detections after the first temperature increase above
20°C were excluded. The first detection above 20°C was retained and marked as
the inferred predation event.


```{r abacus-plot, fig.width=14, fig.height=8, fig.cap="Detection histories and recorded transmitter temperatures of Atlantic salmon kelts released in April (A) and December (B). Each row represents one detected kelt, and points represent individual detections coloured by temperature. Crosses and adjacent values identify the first temperature record above 20°C interpreted as an inferred predation event. Detections following inferred predation were excluded."}

# Add cohort information to detections
det_plot <- det %>%
  left_join(
    meta %>% select(fish, cohort),
    by = "fish"
  ) %>%
  left_join(
    predation_events %>% select(fish, predation_dt),
    by = "fish"
  ) %>%
  filter(is.na(predation_dt) | dt <= predation_dt) %>%
  mutate(
    detection_date = as.Date(dt),
    fish = factor(fish)
  ) %>%
  # Cooler detections are drawn first and warmer detections are drawn on top
  arrange(is.na(temp), temp)

# Prepare predation-event labels
predation_labels <- predation_events %>%
  left_join(
    meta %>% select(fish, cohort),
    by = "fish"
  ) %>%
  mutate(
    detection_date = as.Date(predation_dt),
    fish = factor(fish),
    temperature_label = paste0(round(predation_temp, 1), "°C")
  )

# Data for each cohort
april_plot_data <- det_plot %>%
  filter(cohort == "April") %>%
  mutate(
    fish = factor(
      fish,
      levels = rev(sort(unique(as.numeric(as.character(fish)))))
    )
  )

december_plot_data <- det_plot %>%
  filter(cohort == "December") %>%
  mutate(
    fish = factor(
      fish,
      levels = rev(sort(unique(as.numeric(as.character(fish)))))
    )
  )

april_predation <- predation_labels %>%
  filter(cohort == "April") %>%
  mutate(
    fish = factor(
      fish,
      levels = levels(april_plot_data$fish)
    )
  )

december_predation <- predation_labels %>%
  filter(cohort == "December") %>%
  mutate(
    fish = factor(
      fish,
      levels = levels(december_plot_data$fish)
    )
  )

# Shared plot formatting
common_theme <- theme_minimal(base_size = 16) +
  theme(
    axis.title.x = element_text(size = 18, margin = margin(t = 12)),
    axis.title.y = element_text(size = 18, margin = margin(r = 12)),
    axis.text.x = element_text(
      size = 14,
      angle = 45,
      hjust = 1
    ),
    axis.text.y = element_text(size = 14),
    panel.grid.major.y = element_line(colour = "grey80"),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank(),
    plot.tag = element_text(face = "bold", size = 18),
    plot.title = element_text(face = "bold", size = 18),
    legend.title = element_text(size = 16),
    legend.text = element_text(size = 14)
  )

temperature_scale <- scale_colour_viridis_c(
  option = "plasma",
  limits = c(0, 40),
  oob = scales::squish,
  name = "Temperature (°C)"
)

# April release cohort
p_april <- ggplot(
  april_plot_data,
  aes(x = detection_date, y = fish, colour = temp)
) +
  geom_point(size = 3, alpha = 0.8, na.rm = TRUE) +
  geom_point(
    data = april_predation,
    aes(x = detection_date, y = fish),
    inherit.aes = FALSE,
    shape = 4,
    size = 4,
    stroke = 1.2,
    colour = "black"
  ) +
  geom_text(
    data = april_predation,
    aes(
      x = detection_date,
      y = fish,
      label = temperature_label
    ),
    inherit.aes = FALSE,
    nudge_x = 4,
    hjust = 0,
    size = 5,
    colour = "black"
  ) +
  temperature_scale +
  scale_x_date(
    limits = as.Date(c("2024-04-11", "2024-08-19")),
    date_breaks = "2 weeks",
    date_labels = "%b %d",
    expand = expansion(mult = c(0.01, 0.08))
  ) +
  labs(
    x = "Date",
    y = "Fish",
    title = "April release",
    tag = "A"
  ) +
  common_theme

# December release cohort
p_december <- ggplot(
  december_plot_data,
  aes(x = detection_date, y = fish, colour = temp)
) +
  geom_point(size = 3, alpha = 0.8, na.rm = TRUE) +
  geom_point(
    data = december_predation,
    aes(x = detection_date, y = fish),
    inherit.aes = FALSE,
    shape = 4,
    size = 4,
    stroke = 1.2,
    colour = "black"
  ) +
  geom_text(
    data = december_predation,
    aes(
      x = detection_date,
      y = fish,
      label = temperature_label
    ),
    inherit.aes = FALSE,
    nudge_x = 8,
    hjust = 0,
    size = 5,
    colour = "black"
  ) +
  temperature_scale +
  scale_x_date(
    limits = as.Date(c("2024-12-07", "2025-11-07")),
    date_breaks = "6 weeks",
    date_labels = "%b %d",
    expand = expansion(mult = c(0.01, 0.08))
  ) +
  labs(
    x = "Date",
    y = "Fish",
    title = "December release",
    tag = "B"
  ) +
  common_theme

# Combine panels and collect the shared legend
abacus_plot <- (p_april | p_december) +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom") &
  guides(
    colour = guide_colourbar(
      barwidth = grid::unit(16, "cm"),
      barheight = grid::unit(0.5, "cm")
    )
  )

abacus_plot

ggsave(
  filename = "abacus_plot.png",
  plot = abacus_plot,
  width = 14,
  height = 8,
  units = "in",
  dpi = 600,
  bg = "white"
)
```

# Firth-penalized Cox proportional hazards model

Only fish detected in the marine receiver array were included because the fate
and observation time of undetected fish could not be established. Time was
measured from release to the first inferred predation event. Fish without an
inferred predation event were right-censored at their last marine detection.
Total length was expressed in 5-cm units so that its hazard ratio represented
the effect of a 5-cm increase in total length.

```{r cox-data}
# Prepare fish-level data for the Cox model
cox_data <- fish_fates %>%
  filter(
    detected,
    !is.na(tl),
    !is.na(sex)
  ) %>%
  mutate(
    # Predated fish end at predation; other fish end at their last detection
    endpoint_dt = if_else(
      predation,
      predation_dt,
      last_detection_dt
    ),

    # Time from release to predation or censoring
    time_days = as.numeric(
      difftime(
        endpoint_dt,
        release_dt,
        units = "days"
      )
    ),

    # Event indicator: 1 = inferred predation; 0 = censored
    event = as.integer(predation),

    # Convert total length from metres to units of 5 cm
    tl_5cm = (tl * 100) / 5
  ) %>%
  filter(
    is.finite(time_days),
    time_days > 0
  ) %>%
  droplevels()

# Confirm that usable observations and predation events remain
stopifnot(
  nrow(cox_data) > 0,
  sum(cox_data$event) > 0
)

# Summarize observations included in the model
cox_summary <- cox_data %>%
  summarise(
    n_fish = n(),
    n_predation = sum(event),
    n_censored = sum(event == 0),
    mean_tl_cm = mean(tl * 100),
    min_tl_cm = min(tl * 100),
    max_tl_cm = max(tl * 100)
  )

knitr::kable(
  cox_summary,
  digits = 1,
  caption = "Summary of Atlantic salmon kelts included in the Cox proportional hazards model."
)
```

```{r fit-cox-model}
# Fit the Firth-penalized Cox proportional hazards model
cox_firth <- coxphf(
  formula = Surv(time_days, event) ~ tl_5cm + sex,
  data = cox_data,
  firth = TRUE,
  pl = TRUE
)

summary(cox_firth)
```

## Model checks

Model assumptions were assessed using a standard Cox proportional hazards model
with the same response and covariates as the Firth-penalized model. The standard
model was used only to calculate diagnostic residuals and was not used for
coefficient estimation or inference.

### Proportional hazards assumption

The proportional hazards assumption was assessed using scaled Schoenfeld
residuals. A non-significant test indicates no evidence that a covariate's
effect changes systematically over the observation period.

```{r proportional-hazards-check, fig.width=9, fig.height=4.5}
# Fit an equivalent standard Cox model for diagnostic purposes
cox_diagnostic <- coxph(
  formula = Surv(time_days, event) ~ tl_5cm + sex,
  data = cox_data,
  x = TRUE
)

# Test the proportional hazards assumption
ph_test <- cox.zph(cox_diagnostic)

ph_test

# Plot scaled Schoenfeld residuals
plot(
  ph_test,
  resid = TRUE,
  se = TRUE
)
```

The test for each covariate evaluates whether its scaled Schoenfeld residuals
are associated with time. The global test evaluates the proportional hazards
assumption for the model as a whole. P values greater than 0.05 indicate that
there is insufficient evidence to reject the proportional hazards assumption.
In the diagnostic plots, an approximately horizontal smoothed line suggests
that the effect of the covariate remains reasonably constant through time.
Wide confidence bands are expected because relatively few inferred predation
events were observed.

### Cox–Snell residuals

Cox–Snell residuals were used to assess overall model fit. If the model
adequately represents the data, the estimated cumulative hazard of the
Cox–Snell residuals should approximately follow a straight line with an
intercept of zero and slope of one.

```{r cox-snell-check, fig.width=6.5, fig.height=6}
# Extract martingale residuals from the diagnostic Cox model
martingale_residuals <- residuals(
  cox_diagnostic,
  type = "martingale"
)

# Cox–Snell residual = event indicator - martingale residual
cox_data_diagnostic <- cox_data %>%
  mutate(
    cox_snell = event - martingale_residuals
  )

# Estimate the cumulative hazard of the Cox–Snell residuals
cox_snell_fit <- survfit(
  Surv(cox_snell, event) ~ 1,
  data = cox_data_diagnostic
)

# Retain finite cumulative-hazard estimates for plotting
cox_snell_plot <- tibble(
  residual = cox_snell_fit$time,
  cumulative_hazard = -log(cox_snell_fit$surv)
) %>%
  filter(
    is.finite(residual),
    is.finite(cumulative_hazard)
  )

# Plot estimated cumulative hazard against Cox–Snell residuals
ggplot(
  cox_snell_plot,
  aes(x = residual, y = cumulative_hazard)
) +
  geom_step(
    linewidth = 1,
    colour = "black"
  ) +
  geom_abline(
    intercept = 0,
    slope = 1,
    linewidth = 1,
    linetype = "dashed",
    colour = "red"
  ) +
  coord_equal() +
  labs(
    x = "Cox–Snell residual",
    y = "Estimated cumulative hazard"
  ) +
  theme_classic(base_size = 14)
```

Close agreement between the estimated cumulative hazard and the dashed
45-degree reference line indicates adequate overall model fit. Some divergence
near the upper end of the residual distribution may occur because relatively
few fish remain under observation at longer follow-up times. Given the limited
number of inferred predation events, these diagnostic plots were interpreted
cautiously.

# Summary Table
```{r}
# Create a summary table for fish with inferred predation events

# Last detection before the inferred predation event
last_alive_detection <- det %>%
  inner_join(
    predation_events %>% select(fish, predation_dt),
    by = "fish"
  ) %>%
  filter(dt < predation_dt) %>%
  group_by(fish) %>%
  slice_max(
    order_by = dt,
    n = 1,
    with_ties = FALSE
  ) %>%
  ungroup() %>%
  select(
    fish,
    last_alive_dt = dt
  )

# Temperature records and detections from the predation event onward
post_predation_summary <- det %>%
  inner_join(
    predation_events %>% select(fish, predation_dt),
    by = "fish"
  ) %>%
  filter(dt >= predation_dt) %>%
  group_by(fish) %>%
  summarise(
    final_post_predation_dt = max(dt, na.rm = TRUE),

    post_predation_min_temp = if_else(
      all(is.na(temp)),
      NA_real_,
      min(temp, na.rm = TRUE)
    ),

    post_predation_max_temp = if_else(
      all(is.na(temp)),
      NA_real_,
      max(temp, na.rm = TRUE)
    ),

    post_predation_mean_temp = if_else(
      all(is.na(temp)),
      NA_real_,
      mean(temp, na.rm = TRUE)
    ),

    .groups = "drop"
  )

# Assemble the final table
predation_table <- predation_events %>%
  select(
    fish,
    predation_dt,
    predation_temp
  ) %>%
  left_join(
    meta %>%
      select(fish, cohort, tl),
    by = "fish"
  ) %>%
  left_join(
    last_alive_detection,
    by = "fish"
  ) %>%
  left_join(
    post_predation_summary,
    by = "fish"
  ) %>%
  mutate(
    # Convert total length from metres to centimetres
    total_length_cm = tl * 100,

    # Interval from the final detection before predation to inferred predation
    days_between_detections = as.numeric(
      difftime(
        predation_dt,
        last_alive_dt,
        units = "days"
      )
    ),

    # Duration over which the tag was detected after inferred predation
    days_detected_after_predation = as.numeric(
      difftime(
        final_post_predation_dt,
        predation_dt,
        units = "days"
      )
    ),

    # Combine the minimum and maximum post-predation temperatures
    post_predation_temp_range = if_else(
      is.na(post_predation_min_temp) |
        is.na(post_predation_max_temp),
      NA_character_,
      paste0(
        round(post_predation_min_temp, 1),
        "–",
        round(post_predation_max_temp, 1)
      )
    )
  ) %>%
  transmute(
    Fish = fish,
    `Release cohort` = as.character(cohort),
    `Total length (cm)` = round(total_length_cm, 1),
    `Last alive detection date` = as.Date(last_alive_dt),
    `Approximate predation date` = as.Date(predation_dt),
    `Days between detections` = round(days_between_detections, 1),
    `Predation temperature (°C)` = round(predation_temp, 1),
    `Temperature range post-predation event (°C)` =
      post_predation_temp_range,
    `Mean post-predation temperature (°C)` =
      round(post_predation_mean_temp, 1),
    `Days detected after approximate predation date` =
      round(days_detected_after_predation, 1)
  ) %>%
  arrange(`Approximate predation date`)

predation_table

knitr::kable(
  predation_table,
  align = "c",
  caption = paste(
    "Detection and temperature characteristics of Atlantic salmon kelts",
    "with inferred predation events."
  )
)
```

```{r}
readr::write_csv(
  predation_table,
  "predation_event_summary.csv",
  na = "NA"
)
```



# Reproducibility

```{r session-info}
sessionInfo()
```
