---
title: "Spatial predation patterns"
author: "Caliyena Brown"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  dpi = 300
)

library(dplyr)
library(tidyr)
library(readr)
library(lubridate)
library(sf)
library(ggplot2)
library(ggspatial)
library(rnaturalearth)
library(scales)
library(svglite)
```

# Purpose

This document assigns approximate locations to the finalized inferred
predation events, constructs overall and month-specific 100% minimum convex
polygons (MCPs), calculates MCP areas, and quantifies the spatial and temporal
separation between the last confirmed alive and first post-predation
detections. It exports clean base maps as PNG and SVG files for final assembly
in Inkscape.

# Analysis settings

```{r settings}
# Coordinate reference systems
geographic_crs <- 4326
utm_crs <- 32620

# Map extent used for the Bay of Fundy analysis and figures
map_xlim <- c(-68.9, -64.5)
map_ylim <- c(43.9, 45.5)

# Months must contain more than this number of events to receive an MCP
minimum_monthly_events <- 2

month_levels <- month.name

month_colours <- c(
  "January" = "#D9D9D9",
  "February" = "#D9D9D9",
  "March" = "#D9D9D9",
  "April" = "#E4A3A9",
  "May" = "#D3D3D3",
  "June" = "#D8576E",
  "July" = "#FFC000",
  "August" = "#4EA72E",
  "September" = "#13501B",
  "October" = "#D9D9D9",
  "November" = "#D9D9D9",
  "December" = "#D9D9D9"
)

```

# Import and prepare data

This file expects the cleaned detections and finalized predation events used in
the mortality and temporal analyses. It also expects an analysis-ready receiver
file containing one row per station in the Bay of Fundy array.

```{r import-data}
det <- readRDS("det_clean_no_release_site_det.rds")
predation_events <- readRDS("predation_events_clean.rds")
receivers <- read_csv(
  "receiver_metadata_clean.csv",
  show_col_types = FALSE
)

required_det_columns <- c(
  "fish", "dt", "lon", "lat", "station", "receiver"
)

required_predation_columns <- c(
  "fish", "predation_dt", "predation_temp"
)

required_receiver_columns <- c(
  "station", "lon", "lat", "marine_farm"
)

stopifnot(
  all(required_det_columns %in% names(det)),
  all(required_predation_columns %in% names(predation_events)),
  all(required_receiver_columns %in% names(receivers))
)

# Confirm that the cleaned receiver file contains valid, unique stations.
receivers <- receivers %>%
  mutate(
    station = as.character(station),
    lon = as.numeric(lon),
    lat = as.numeric(lat),
    marine_farm = as.logical(marine_farm)
  ) %>%
  filter(
    !is.na(station),
    !is.na(lon),
    !is.na(lat)
  ) %>%
  distinct(station, .keep_all = TRUE)

if (anyNA(receivers$marine_farm)) {
  stop("The marine_farm column must contain only TRUE or FALSE values.")
}

marine_farm_receivers <- receivers %>%
  filter(marine_farm)

# Convert detection and event times to UTC without discarding time of day.
if (inherits(det$dt, "POSIXct")) {
  det$dt <- with_tz(det$dt, tzone = "UTC")
} else {
  det$dt <- ymd_hms(det$dt, tz = "UTC", quiet = TRUE)
}

if (inherits(predation_events$predation_dt, "POSIXct")) {
  predation_events$predation_dt <- with_tz(
    predation_events$predation_dt,
    tzone = "UTC"
  )
} else {
  predation_events$predation_dt <- ymd_hms(
    predation_events$predation_dt,
    tz = "UTC",
    quiet = TRUE
  )
}

det <- det %>%
  mutate(
    fish = as.integer(fish),
    lon = as.numeric(lon),
    lat = as.numeric(lat)
  ) %>%
  filter(
    !is.na(fish),
    !is.na(dt)
  ) %>%
  arrange(fish, dt)

predation_events <- predation_events %>%
  mutate(
    fish = as.integer(fish),
    month = factor(
      month.name[month(predation_dt)],
      levels = month_levels,
      ordered = TRUE
    )
  ) %>%
  distinct(fish, .keep_all = TRUE)

stopifnot(
  nrow(predation_events) == 13,
  anyDuplicated(predation_events$fish) == 0
)
```

# Assign approximate predation locations

The approximate predation location is the receiver associated with the saved
first post-predation detection. Matching uses fish identity and the complete
timestamp so that the spatial and temporal analyses use the same events.

```{r assign-event-locations}
predation_locations <- predation_events %>%
  select(
    fish,
    predation_dt,
    predation_temp,
    month
  ) %>%
  left_join(
    det %>%
      filter(
        !is.na(lon),
        !is.na(lat)
      ) %>%
      select(
        fish,
        detection_dt = dt,
        station,
        receiver,
        lon,
        lat
      ) %>%
      distinct(
        fish,
        detection_dt,
        .keep_all = TRUE
      ),
    by = c(
      "fish",
      "predation_dt" = "detection_dt"
    )
  ) %>%
  filter(
    !is.na(lon),
    !is.na(lat),
    station %in% receivers$station,
    between(lon, map_xlim[1], map_xlim[2]),
    between(lat, map_ylim[1], map_ylim[2])
  ) %>%
  arrange(predation_dt)

if (nrow(predation_locations) != nrow(predation_events)) {
  missing_fish <- setdiff(
    predation_events$fish,
    predation_locations$fish
  )

  stop(
    "Predation locations could not be assigned for fish: ",
    paste(missing_fish, collapse = ", ")
  )
}

predation_locations_sf <- st_as_sf(
  predation_locations,
  coords = c("lon", "lat"),
  crs = geographic_crs,
  remove = FALSE
)

predation_locations_utm <- st_transform(
  predation_locations_sf,
  utm_crs
)

predation_locations
```

# Construct overall and monthly MCPs

```{r construct-mcps}
# Confirm that at least three unique locations are available for the overall MCP.
if (nrow(distinct(predation_locations, lon, lat)) < 3) {
  stop("At least three unique predation locations are required for an MCP.")
}

# Construct the 100% MCP from all assigned predation locations.
mcp_all_utm <- predation_locations_utm %>%
  summarise(geometry = st_convex_hull(st_union(geometry)))

mcp_all_ll <- st_transform(mcp_all_utm, geographic_crs)

mcp_all_area <- mcp_all_utm %>%
  mutate(
    group = "All events",
    n_events = nrow(predation_locations),
    area_km2 = as.numeric(st_area(geometry)) / 1e6
  ) %>%
  st_drop_geometry() %>%
  select(group, n_events, area_km2)

# Identify months containing more than two predation events and at least three
# unique coordinates.
months_for_mcp <- predation_locations %>%
  group_by(month) %>%
  summarise(
    n_events = n(),
    n_locations = n_distinct(paste(lon, lat)),
    .groups = "drop"
  ) %>%
  filter(
    n_events > minimum_monthly_events,
    n_locations >= 3
  )

# Construct one 100% MCP for every qualifying month.
mcp_month_utm <- predation_locations_utm %>%
  filter(month %in% months_for_mcp$month) %>%
  group_by(month) %>%
  summarise(
    n_events = n(),
    geometry = st_convex_hull(st_union(geometry)),
    .groups = "drop"
  )

mcp_month_ll <- st_transform(mcp_month_utm, geographic_crs)

mcp_month_areas <- mcp_month_utm %>%
  mutate(
    area_km2 = as.numeric(st_area(geometry)) / 1e6
  ) %>%
  st_drop_geometry() %>%
  arrange(month)

mcp_area_summary <- bind_rows(
  mcp_all_area,
  mcp_month_areas %>%
    transmute(
      group = as.character(month),
      n_events,
      area_km2
    )
)

knitr::kable(
  mcp_area_summary,
  digits = 1,
  caption = "Areas encompassed by overall and month-specific 100% MCPs."
)

write_csv(
  mcp_area_summary,
  "predation_mcp_area_summary.csv",
  na = "NA"
)

# Export spatial objects for inspection or reuse.
st_write(
  mcp_all_ll,
  "predation_mcp_overall.gpkg",
  layer = "overall_mcp",
  delete_dsn = TRUE,
  quiet = TRUE
)

st_write(
  mcp_month_ll,
  "predation_mcp_monthly.gpkg",
  layer = "monthly_mcps",
  delete_dsn = TRUE,
  quiet = TRUE
)
```

# Quantify spatial and temporal uncertainty

```{r spatial-uncertainty-data}
# Identify the immediately preceding detection with valid coordinates.
last_alive_locations <- det %>%
  inner_join(
    predation_locations %>%
      select(fish, predation_dt),
    by = "fish"
  ) %>%
  filter(
    dt < predation_dt,
    !is.na(lon),
    !is.na(lat)
  ) %>%
  group_by(fish) %>%
  slice_max(
    order_by = dt,
    n = 1,
    with_ties = FALSE
  ) %>%
  ungroup() %>%
  transmute(
    fish,
    last_alive_dt = dt,
    last_alive_station = station,
    last_alive_lon = lon,
    last_alive_lat = lat
  )

uncertainty_pairs <- predation_locations %>%
  transmute(
    fish,
    predation_dt,
    predation_station = station,
    predation_lon = lon,
    predation_lat = lat,
    predation_temp,
    month
  ) %>%
  left_join(last_alive_locations, by = "fish") %>%
  mutate(
    interval_hours = as.numeric(
      difftime(
        predation_dt,
        last_alive_dt,
        units = "hours"
      )
    ),
    interval_days = interval_hours / 24
  )

if (anyNA(uncertainty_pairs$last_alive_lon) ||
    anyNA(uncertainty_pairs$last_alive_lat)) {
  stop("At least one fish is missing a preceding detection location.")
}

# Convert paired coordinates to projected points.
last_alive_sf <- st_as_sf(
  uncertainty_pairs,
  coords = c("last_alive_lon", "last_alive_lat"),
  crs = geographic_crs,
  remove = FALSE
) %>%
  st_transform(utm_crs)

post_predation_sf <- st_as_sf(
  uncertainty_pairs,
  coords = c("predation_lon", "predation_lat"),
  crs = geographic_crs,
  remove = FALSE
) %>%
  st_transform(utm_crs)

# Calculate straight-line distance between paired receiver locations.
spatial_distance_km <- as.numeric(
  st_distance(
    st_geometry(last_alive_sf),
    st_geometry(post_predation_sf),
    by_element = TRUE
  )
) / 1000

uncertainty_summary <- uncertainty_pairs %>%
  mutate(
    spatial_distance_km = spatial_distance_km
  ) %>%
  select(
    fish,
    month,
    last_alive_dt,
    predation_dt,
    interval_hours,
    interval_days,
    last_alive_station,
    predation_station,
    last_alive_lon,
    last_alive_lat,
    predation_lon,
    predation_lat,
    spatial_distance_km,
    predation_temp
  ) %>%
  arrange(desc(spatial_distance_km))

knitr::kable(
  uncertainty_summary,
  digits = 2,
  caption = paste(
    "Temporal and spatial separation between the last confirmed alive",
    "and first post-predation detections."
  )
)

write_csv(
  uncertainty_summary,
  "predation_spatial_uncertainty.csv",
  na = "NA"
)
```

# Prepare map layers

```{r map-layers}
# Combine Canada and the United States for the coastline basemap.
land <- rnaturalearth::ne_countries(
  scale = "medium",
  returnclass = "sf"
) %>%
  filter(admin %in% c("Canada", "United States of America")) %>%
  st_transform(geographic_crs)

receivers_sf <- st_as_sf(
  receivers,
  coords = c("lon", "lat"),
  crs = geographic_crs,
  remove = FALSE
)

marine_farm_sf <- st_as_sf(
  marine_farm_receivers,
  coords = c("lon", "lat"),
  crs = geographic_crs,
  remove = FALSE
)

# Build one line segment from the last alive to post-predation location for
# each fish. Lines are constructed in UTM and transformed back for plotting.
segment_list <- lapply(seq_len(nrow(uncertainty_pairs)), function(i) {
  coordinates <- matrix(
    c(
      uncertainty_pairs$last_alive_lon[i],
      uncertainty_pairs$last_alive_lat[i],
      uncertainty_pairs$predation_lon[i],
      uncertainty_pairs$predation_lat[i]
    ),
    ncol = 2,
    byrow = TRUE
  )

  st_linestring(coordinates)
})

uncertainty_segments <- st_sf(
  fish = uncertainty_pairs$fish,
  geometry = st_sfc(segment_list, crs = geographic_crs)
)

last_alive_plot_sf <- st_as_sf(
  uncertainty_pairs,
  coords = c("last_alive_lon", "last_alive_lat"),
  crs = geographic_crs,
  remove = FALSE
)

post_predation_plot_sf <- st_as_sf(
  uncertainty_pairs,
  coords = c("predation_lon", "predation_lat"),
  crs = geographic_crs,
  remove = FALSE
)

fish_levels <- sort(unique(uncertainty_pairs$fish))

shape_values <- c(
  0, 3, 18, 17, 8, 5, 6,
  1, 15, 16, 2, 4, 7
)

if (length(fish_levels) > length(shape_values)) {
  stop("Add more point shapes for the number of fish being plotted.")
}

fish_shapes <- setNames(
  shape_values[seq_along(fish_levels)],
  as.character(fish_levels)
)

last_alive_plot_sf <- last_alive_plot_sf %>%
  mutate(
    fish_id = factor(fish, levels = fish_levels)
  )

post_predation_plot_sf <- post_predation_plot_sf %>%
  mutate(
    fish_id = factor(fish, levels = fish_levels)
  )
```

# Panel A: predation extent and monthly MCPs

```{r panel-a, fig.width=10, fig.height=6.5, fig.cap="Approximate locations and 100% MCPs of inferred predation events in the Bay of Fundy. The light-blue polygon represents the overall MCP; coloured polygons represent months containing more than two events."}
panel_a <- ggplot() +
  # Overall and monthly MCPs are drawn before land so land masks polygon areas
  # that cross coastlines without changing the calculated MCP areas.
  geom_sf(
    data = mcp_all_ll,
    fill = "lightblue",
    colour = "midnightblue",
    alpha = 0.35,
    linewidth = 0.5
  ) +
  geom_sf(
    data = mcp_month_ll,
    aes(fill = month, colour = month),
    alpha = 0.30,
    linewidth = 0.7
  ) +
  geom_sf(
    data = land,
    fill = "grey97",
    colour = "grey70",
    linewidth = 0.3
  ) +
  geom_sf(
    data = receivers_sf,
    shape = 16,
    colour = "grey45",
    size = 1,
    alpha = 0.65
  ) +
  geom_sf(
    data = predation_locations_sf,
    aes(colour = month),
    size = 3.2,
    alpha = 0.95
  ) +
  scale_colour_manual(
    values = month_colours,
    drop = TRUE,
    name = "Month"
  ) +
  scale_fill_manual(
    values = month_colours,
    drop = TRUE,
    name = "Month"
  ) +
  coord_sf(
    xlim = map_xlim,
    ylim = map_ylim,
    expand = FALSE
  ) +
  annotation_scale(
    location = "br",
    width_hint = 0.18
  ) +
  annotation_north_arrow(
    location = "br",
    which_north = "true",
    pad_x = grid::unit(0.4, "cm"),
    pad_y = grid::unit(0.9, "cm"),
    style = north_arrow_fancy_orienteering,
    height = grid::unit(0.9, "cm"),
    width = grid::unit(0.8, "cm")
  ) +
  labs(
    x = "Longitude",
    y = "Latitude"
  ) +
  theme_classic(base_size = 13) +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 13),
    legend.text = element_text(size = 11)
  )

panel_a

ggsave(
  "spatial_predation_panel_A.png",
  panel_a,
  width = 10,
  height = 6.5,
  units = "in",
  dpi = 600,
  bg = "white"
)

ggsave(
  "spatial_predation_panel_A.svg",
  panel_a,
  width = 10,
  height = 6.5,
  units = "in",
  bg = "white"
)
```

# Panel B: spatial uncertainty

```{r panel-b, fig.width=10, fig.height=6.5, fig.cap="Spatial uncertainty between the last confirmed alive and first post-predation detections. Red and blue symbols show paired detections for each fish; black lines connect the locations. Gold diamonds indicate marine-farm receiver stations."}
panel_b <- ggplot() +
  geom_sf(
    data = mcp_all_ll,
    fill = "lightblue",
    colour = "lightblue",
    alpha = 0.35,
    linewidth = 0.4
  ) +
  geom_sf(
    data = land,
    fill = "grey97",
    colour = "grey70",
    linewidth = 0.3
  ) +
  geom_sf(
    data = marine_farm_sf,
    shape = 23,
    colour = "black",
    fill = "gold",
    stroke = 0.4,
    size = 2.8,
    alpha = 0.8
  ) +
  geom_sf(
    data = uncertainty_segments,
    colour = "black",
    linewidth = 0.5,
    alpha = 0.8
  ) +
  geom_sf(
    data = last_alive_plot_sf,
    aes(shape = fish_id),
    colour = "red3",
    size = 3.5,
    stroke = 0.9
  ) +
  geom_sf(
    data = post_predation_plot_sf,
    aes(shape = fish_id),
    colour = "midnightblue",
    size = 3.5,
    stroke = 0.9
  ) +
  scale_shape_manual(
    values = fish_shapes,
    name = "Fish ID"
  ) +
  coord_sf(
    xlim = map_xlim,
    ylim = map_ylim,
    expand = FALSE
  ) +
  annotation_scale(
    location = "br",
    width_hint = 0.18
  ) +
  annotation_north_arrow(
    location = "br",
    which_north = "true",
    pad_x = grid::unit(0.4, "cm"),
    pad_y = grid::unit(0.9, "cm"),
    style = north_arrow_fancy_orienteering,
    height = grid::unit(0.9, "cm"),
    width = grid::unit(0.8, "cm")
  ) +
  labs(
    x = "Longitude",
    y = "Latitude"
  ) +
  theme_classic(base_size = 13) +
  theme(
    legend.position = "right",
    legend.title = element_text(size = 13),
    legend.text = element_text(size = 10)
  )

panel_b

ggsave(
  "spatial_predation_panel_B.png",
  panel_b,
  width = 10,
  height = 6.5,
  units = "in",
  dpi = 600,
  bg = "white"
)

ggsave(
  "spatial_predation_panel_B.svg",
  panel_b,
  width = 10,
  height = 6.5,
  units = "in",
  bg = "white"
)
```

# Reproducibility

```{r session-info}
sessionInfo()
```
