---
title: "Temporal habitat use of inner Bay of Fundy Atlantic salmon kelts"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
date: "2026-08-18"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  dpi = 300
)

library(dplyr)
library(tidyr)
library(readr)
library(lubridate)
library(ggplot2)
library(mgcv)
library(sf)
library(scales)
```

# Purpose

This analysis evaluates monthly spatial patterns in the presence of tagged inner
Bay of Fundy Atlantic salmon kelts. Detections are truncated at each inferred
predation event, aggregated to receiver station by year-month, and modelled using a
binomial generalized additive model (GAM). Month-specific spatial smooths allow
the spatial pattern of receiver use to vary among calendar months.

# Analysis settings

```{r settings}
crs_geographic <- 4326
crs_projected <- 32620
minimum_fish_per_month <- 5L
initial_basis_dimension <- 30L
candidate_basis_dimension <- 50L
prediction_resolution_m <- 1000

# Final month included in the study period
study_end_month <- as.Date("2025-10-01")

month_labels <- setNames(month.name, 1:12)
```

# Import clean data

The analysis uses the cleaned detections, finalized predation events, biological
metadata, and an analysis-ready receiver table. The receiver table contains one 
row per station in the primary Bay of Fundy receiver grid, with station 
coordinates recorded as decimal degrees. Stations outside the primary receiver 
grid were removed when preparing this file.

Required receiver columns are `station`, `lon`, and `lat`.

```{r import-data}
det <- readRDS("det_clean_no_release_site_det.rds")
predation_events <- readRDS("predation_events_clean.rds")
meta <- readRDS("meta_clean.rds")

receivers <- read_csv(
  "recs_meta.csv",
  show_col_types = FALSE
)

# Polygon layer containing DFO marine habitat areas 3–8
gdb_path <- "IBOF_MarineCriticalHabitats.gdb"

habitat_areas <- st_read(
  gdb_path,
  layer = "IBOF_Crit_Hab_Polygons_EqAr",
  quiet = TRUE
) %>%
  filter(CH_UNIT_ID %in% 3:8)

stopifnot(
  nrow(habitat_areas) == 6,
  all(sort(habitat_areas$CH_UNIT_ID) == 3:8)
)

required_det_columns <- c(
  "fish", "dt", "receiver", "station", "lon", "lat"
)

required_predation_columns <- c(
  "fish", "predation_dt"
)

required_meta_columns <- c(
  "fish", "release_dt"
)

required_receiver_columns <- c(
  "station", "lon", "lat"
)

stopifnot(
  all(required_det_columns %in% names(det)),
  all(required_predation_columns %in% names(predation_events)),
  all(required_meta_columns %in% names(meta)),
  all(required_receiver_columns %in% names(receivers))
)
```

# Prepare detections and receiver stations

```{r prepare-data}
# Convert date-time fields to UTC
parse_utc <- function(x) {
  if (inherits(x, "POSIXt")) {
    return(with_tz(x, tzone = "UTC"))
  }

  if (inherits(x, "Date")) {
    return(as.POSIXct(x, tz = "UTC"))
  }

  parse_date_time(
    x,
    orders = c(
      "ymd HMS",
      "ymd HM",
      "ymd"
    ),
    tz = "UTC",
    quiet = TRUE
  )
}

# Prepare detection data
det <- det %>%
  mutate(
    fish = as.integer(fish),
    station = as.character(station),
    receiver = as.character(receiver),
    dt = parse_utc(dt),
    lon = as.numeric(lon),
    lat = as.numeric(lat)
  ) %>%
  filter(
    !is.na(fish),
    !is.na(station),
    !is.na(dt)
  ) %>%
  arrange(fish, dt)

# Prepare finalized inferred predation events
predation_events <- predation_events %>%
  transmute(
    fish = as.integer(fish),
    predation_dt = parse_utc(predation_dt)
  ) %>%
  distinct(fish, .keep_all = TRUE)

# Prepare biological metadata
meta <- meta %>%
  transmute(
    fish = as.integer(fish),
    release_dt = parse_utc(release_dt)
  ) %>%
  distinct(fish, .keep_all = TRUE) %>%
  mutate(
    cohort = case_when(
      month(release_dt) == 4 ~ "April",
      month(release_dt) == 12 ~ "December",
      TRUE ~ NA_character_
    ),
    cohort = factor(
      cohort,
      levels = c("April", "December")
    )
  )

stopifnot(!anyNA(meta$cohort))

# Prepare the receiver stations in the primary receiver grid
receivers <- receivers %>%
  transmute(
    station = as.character(station),
    lon = as.numeric(lon),
    lat = as.numeric(lat)
  ) %>%
  filter(
    !is.na(station),
    !is.na(lon),
    !is.na(lat)
  ) %>%
  distinct(station, .keep_all = TRUE)

# Confirm that receiver stations and predation events are unique
stopifnot(
  anyDuplicated(receivers$station) == 0,
  nrow(predation_events) == 13,
  anyDuplicated(predation_events$fish) == 0,
  anyDuplicated(meta$fish) == 0
)

# Exclude detections after each fish's inferred predation event
det_analysis <- det %>%
  left_join(
    predation_events,
    by = "fish"
  ) %>%
  filter(
    is.na(predation_dt) |
      dt <= predation_dt
  ) %>%
  select(-predation_dt)

# Retain only stations in the cleaned primary receiver grid
det_analysis <- det_analysis %>%
  semi_join(
    receivers %>% select(station),
    by = "station"
  ) %>%
  mutate(
    ym = floor_date(dt, unit = "month"),
    month = month(ym)
  )

stopifnot(nrow(det_analysis) > 0)
```

# Select calendar months

Analyses are restricted to calendar months in which at least five unique kelts
were detected within the primary receiver grid after applying the predation
truncation.

```{r select-months}
fish_month_count <- det_analysis %>%
  group_by(month) %>%
  summarise(
    n_fish = n_distinct(fish),
    n_detections = n(),
    .groups = "drop"
  ) %>%
  arrange(month) %>%
  mutate(
    month_name = month.name[month]
  )

months_keep <- fish_month_count %>%
  filter(
    n_fish >= minimum_fish_per_month
  ) %>%
  pull(month) %>%
  sort()

stopifnot(length(months_keep) >= 2)

knitr::kable(
  fish_month_count,
  caption = paste(
    "Numbers of unique tagged kelts and detections by",
    "calendar month within the primary receiver grid."
  )
)

fish_by_month_plot <- fish_month_count %>%
  mutate(
    month_name = factor(
      month_name,
      levels = month.name
    )
  ) %>%
  ggplot(
    aes(
      x = month_name,
      y = n_fish
    )
  ) +
  geom_col(
    fill = "grey55",
    colour = "black",
    linewidth = 0.3
  ) +
  geom_hline(
    yintercept = minimum_fish_per_month,
    linetype = "dashed"
  ) +
  labs(
    x = "Month",
    y = "Number of unique kelts detected"
  ) +
  theme_bw(base_size = 12) +
  theme(
    axis.text.x = element_text(
      angle = 45,
      hjust = 1
    )
  )

fish_by_month_plot

ggsave(
  filename = "gam_fish_detected_by_month.png",
  plot = fish_by_month_plot,
  width = 8,
  height = 5.5,
  units = "in",
  dpi = 600,
  bg = "white"
)
```
# Summarize detections by year and release cohort

Detections from the same calendar month were pooled across study years and
release cohorts for the GAM. The following tables show how each release cohort
and study year contributed to the pooled monthly data.

```{r monthly-cohort-summary}
# Add release cohort to the detection data
detections_by_cohort <- det_analysis %>%
  left_join(
    meta %>%
      select(
        fish,
        cohort
      ),
    by = "fish"
  ) %>%
  mutate(
    detection_year = year(dt),
    calendar_month = month(dt)
  )

stopifnot(!anyNA(detections_by_cohort$cohort))

# Count unique fish, stations, and detections
monthly_cohort_counts <- detections_by_cohort %>%
  group_by(
    detection_year,
    calendar_month,
    cohort
  ) %>%
  summarise(
    n_unique_fish = n_distinct(fish),
    n_stations = n_distinct(station),
    n_detections = n(),
    .groups = "drop"
  )

# Include combinations with no detections
monthly_cohort_summary <- monthly_cohort_counts %>%
  complete(
    detection_year,
    calendar_month = 1:12,
    cohort,
    fill = list(
      n_unique_fish = 0L,
      n_stations = 0L,
      n_detections = 0L
    )
  ) %>%
  mutate(
    month_name = month.name[calendar_month]
  ) %>%
  arrange(
    detection_year,
    calendar_month,
    cohort
  ) %>%
  select(
    detection_year,
    calendar_month,
    month_name,
    cohort,
    n_unique_fish,
    n_stations,
    n_detections
  )

knitr::kable(
  monthly_cohort_summary,
  caption = paste(
    "Numbers of unique kelts, receiver stations, and detections",
    "by year, calendar month, and release cohort."
  )
)

write_csv(
  monthly_cohort_summary,
  "gam_monthly_cohort_detection_summary.csv",
  na = "NA"
)
```

# Check for April-cohort detections during 2025

```{r april-cohort-2025}
april_cohort_2025 <- detections_by_cohort %>%
  filter(
    cohort == "April",
    detection_year == 2025
  ) %>%
  summarise(
    n_unique_fish = n_distinct(fish),
    fish_ids = if_else(
      n_unique_fish > 0,
      paste(
        sort(unique(fish)),
        collapse = ", "
      ),
      "None"
    )
  )

knitr::kable(
  april_cohort_2025,
  caption = "April-release cohort detections recorded during 2025."
)
```

# Visualize monthly sample sizes by cohort and year

```{r monthly-cohort-figure, fig.width=11, fig.height=6}
monthly_cohort_plot_data <- monthly_cohort_summary %>%
  mutate(
    month_name = factor(
      month_name,
      levels = month.name
    )
  )

monthly_cohort_plot <- ggplot(
  monthly_cohort_plot_data,
  aes(
    x = month_name,
    y = n_unique_fish,
    fill = cohort
  )
) +
  geom_col(
    position = "dodge",
    colour = "black",
    linewidth = 0.25
  ) +
  geom_hline(
    yintercept = minimum_fish_per_month,
    linetype = "dashed",
    colour = "black"
  ) +
  facet_wrap(
    ~ detection_year,
    ncol = 1
  ) +
  scale_fill_manual(
    values = c(
      "April" = "#E76F51",
      "December" = "#457B9D"
    ),
    name = "Release cohort"
  ) +
  labs(
    x = "Calendar month",
    y = "Number of unique kelts detected"
  ) +
  theme_bw(base_size = 13) +
  theme(
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(
      angle = 45,
      hjust = 1
    ),
    strip.background = element_blank(),
    strip.text = element_text(
      face = "bold",
      size = 13
    ),
    legend.position = "bottom"
  )

monthly_cohort_plot

ggsave(
  filename = "gam_monthly_fish_by_cohort_and_year.png",
  plot = monthly_cohort_plot,
  width = 11,
  height = 6,
  units = "in",
  dpi = 600,
  bg = "white"
)
```

# Build receiver-station by year-month observations

The presence–absence dataset includes all stations in the cleaned primary
receiver grid. Each station is treated as available for observation throughout
the study period because receiver coverage within this grid was considered
consistent. Presence indicates that at least one tagged kelt was detected at a
station during a given year-month.

```{r receiver-month-data}
# Convert receiver-station coordinates to UTM Zone 20N
receiver_stations_sf <- receivers %>%
  st_as_sf(
    coords = c("lon", "lat"),
    crs = crs_geographic,
    remove = FALSE
  ) %>%
  st_transform(crs_projected)

station_coordinates <- st_coordinates(
  receiver_stations_sf
)

receiver_stations <- receiver_stations_sf %>%
  st_drop_geometry() %>%
  mutate(
    x = station_coordinates[, 1],
    y = station_coordinates[, 2]
  )

# Confirm that all retained detection stations have receiver metadata
unmatched_detection_stations <- det_analysis %>%
  distinct(station) %>%
  anti_join(
    receiver_stations %>%
      select(station),
    by = "station"
  )

stopifnot(nrow(unmatched_detection_stations) == 0)

# Define the study period
first_release_month <- floor_date(
  min(meta$release_dt, na.rm = TRUE),
  unit = "month"
)

study_end_datetime <- as.POSIXct(
  study_end_month,
  tz = "UTC"
)

stopifnot(study_end_datetime >= first_release_month)

# Create the sequence of year-months represented in the analysis
ym_sequence <- seq(
  from = first_release_month,
  to = study_end_datetime,
  by = "1 month"
)

# Retain only calendar months meeting the five-fish criterion
ym_sequence <- ym_sequence[
  month(ym_sequence) %in% months_keep
]

# Create the complete station-by-year-month grid
receiver_effort <- expand_grid(
  station = receiver_stations$station,
  ym = ym_sequence
) %>%
  left_join(
    receiver_stations,
    by = "station"
  ) %>%
  mutate(
    month = month(ym)
  )

# Summarize observed detections to station-by-year-month presence
receiver_month_observed <- det_analysis %>%
  filter(
    month %in% months_keep
  ) %>%
  group_by(
    station,
    ym
  ) %>%
  summarise(
    present = 1L,
    n_detections = n(),
    n_fish = n_distinct(fish),
    .groups = "drop"
  )

# Join observations to the complete grid and fill absences
receiver_month_presence <- receiver_effort %>%
  left_join(
    receiver_month_observed,
    by = c(
      "station",
      "ym"
    )
  ) %>%
  mutate(
    present = replace_na(present, 0L),
    n_detections = replace_na(n_detections, 0L),
    n_fish = replace_na(n_fish, 0L),
    month_f = factor(
      month,
      levels = months_keep
    )
  )

stopifnot(
  nrow(receiver_month_presence) > 0,
  all(receiver_month_presence$present %in% c(0L, 1L)),
  !anyNA(receiver_month_presence$x),
  !anyNA(receiver_month_presence$y),
  !anyNA(receiver_month_presence$month_f)
)

# Summarize the model dataset by calendar month
receiver_month_summary <- receiver_month_presence %>%
  group_by(month_f) %>%
  summarise(
    n_station_months = n(),
    n_stations = n_distinct(station),
    n_presences = sum(present),
    observed_presence_rate = mean(present),
    .groups = "drop"
  )

knitr::kable(
  receiver_month_summary,
  digits = 3,
  caption = paste(
    "Receiver-station by year-month observations used",
    "in the binomial generalized additive model."
  )
)
```

# Fit the binomial GAM (attempt 1 with K = 30)

Calendar month is included as a categorical fixed effect. A separate
two-dimensional thin-plate regression spline is estimated for each retained
month. The default factor contrasts express parametric month coefficients
relative to the first retained month.

```{r fit-k30}
gam_k30 <- bam(
  present ~ month_f +
    s(
      x,
      y,
      by = month_f,
      k = initial_basis_dimension,
      bs = "tp"
    ),
  family = binomial(link = "logit"),
  method = "fREML",
  discrete = TRUE,
  select = TRUE,
  data = receiver_month_presence
)

summary(gam_k30)
anova(gam_k30)
```

# Initial model estimates

Odds ratios below describe parametric differences from the reference month.
They do not summarize the month-specific spatial smooths.

```{r month-odds-ratios}
parametric_table_k30 <- as.data.frame(summary(gam_k30)$p.table) %>%
  tibble::rownames_to_column("term") %>%
  rename(
    estimate = Estimate,
    standard_error = `Std. Error`,
    statistic = `z value`,
    p_value = `Pr(>|z|)`
  ) %>%
  mutate(
    odds_ratio = exp(estimate),
    odds_ratio_lower = exp(estimate - 1.96 * standard_error),
    odds_ratio_upper = exp(estimate + 1.96 * standard_error)
  )

knitr::kable(
  parametric_table_k30,
  digits = 3,
  caption = "Parametric coefficients and odds ratios for the initial k = 30 model."
)
```

# Initial model checks

For a binomial model, raw residuals are not expected to follow a normal
distribution. Diagnostics therefore focus on basis dimensions, concurvity,
calibration, and residual autocorrelation rather than residual normality.

```{r check-k30, fig.width=10, fig.height=8}
set.seed(123)
gam.check(gam_k30)

concurvity_results <- concurvity(
  gam_k30,
  full = TRUE
)

concurvity_results

acf(
  residuals(gam_k30, type = "deviance"),
  main = "ACF of deviance residuals: k = 30"
)
```

```{r calibration-k30, fig.width=11, fig.height=7}
calibration_data_k30 <- receiver_month_presence %>%
  mutate(
    predicted_probability = predict(
      gam_k30,
      type = "response"
    ),
    probability_bin = cut(
      predicted_probability,
      breaks = seq(0, 1, by = 0.1),
      include.lowest = TRUE
    )
  ) %>%
  group_by(month_f, probability_bin) %>%
  summarise(
    observed_rate = mean(present),
    predicted_rate = mean(predicted_probability),
    n = n(),
    .groups = "drop"
  )

calibration_plot_k30 <- ggplot(
  calibration_data_k30,
  aes(
    x = predicted_rate,
    y = observed_rate,
    size = n
  )
) +
  geom_point(alpha = 0.75) +
  geom_abline(
    intercept = 0,
    slope = 1,
    linetype = "dashed",
    colour = "red"
  ) +
  facet_wrap(
    ~ month_f,
    labeller = as_labeller(month_labels)
  ) +
  coord_equal(xlim = c(0, 1), ylim = c(0, 1)) +
  labs(
    x = "Mean predicted probability",
    y = "Observed presence rate",
    size = "Station-months"
  ) +
  theme_bw(base_size = 12)

calibration_plot_k30
```

# Fit and check the k = 50 candidate model

```{r fit-k50}
gam_k50 <- bam(
  present ~ month_f +
    s(
      x,
      y,
      by = month_f,
      k = candidate_basis_dimension,
      bs = "tp"
    ),
  family = binomial(link = "logit"),
  method = "fREML",
  discrete = TRUE,
  select = TRUE,
  data = receiver_month_presence
)

summary(gam_k50)
anova(gam_k50)
```

```{r check-k50, fig.width=10, fig.height=8}
set.seed(123)
gam.check(gam_k50)
```

# Compare predictions from the two models 

```{r compare-predictions}
k_comparison <- receiver_month_presence %>%
  mutate(
    prediction_k30 = predict(
      gam_k30,
      type = "response"
    ),
    prediction_k50 = predict(
      gam_k50,
      type = "response"
    ),
    absolute_difference = abs(
      prediction_k50 - prediction_k30
    )
  )

prediction_comparison <- k_comparison %>%
  summarise(
    prediction_correlation = cor(
      prediction_k30,
      prediction_k50
    ),
    mean_absolute_difference = mean(
      absolute_difference
    ),
    maximum_absolute_difference = max(
      absolute_difference
    )
  )

knitr::kable(
  prediction_comparison,
  digits = 4,
  caption = "Comparison of fitted probabilities from the candidate GAMs."
)
```

# Compare model summaries 

```{r compare-model-fit}
model_comparison <- tibble(
  model = c("k = 30", "k = 50"),
  deviance_explained = c(
    summary(gam_k30)$dev.expl,
    summary(gam_k50)$dev.expl
  ),
  adjusted_r_squared = c(
    summary(gam_k30)$r.sq,
    summary(gam_k50)$r.sq
  ),
  fREML_score = c(
    gam_k30$gcv.ubre,
    gam_k50$gcv.ubre
  )
)

knitr::kable(
  model_comparison,
  digits = 4,
  caption = "Fit statistics for the candidate basis dimensions."
)
```

A basis dimension of `k = 50` was selected following comparison with the initial
`k = 30` model. The larger basis dimension modestly improved model fit and
allowed additional spatial complexity, although predictions from the two models
remained strongly correlated. All subsequent estimates and spatial predictions
use the selected `k = 50` model.

```{r select-final-model}
gam_presence <- gam_k50
```

# Final month effects and odds ratios

Odds ratios describe parametric differences from the first retained calendar
month. They do not summarize the month-specific spatial smooths.

```{r final-month-odds-ratios}
parametric_table <- as.data.frame(summary(gam_presence)$p.table) %>%
  tibble::rownames_to_column("term") %>%
  rename(
    estimate = Estimate,
    standard_error = `Std. Error`,
    statistic = `z value`,
    p_value = `Pr(>|z|)`
  ) %>%
  mutate(
    odds_ratio = exp(estimate),
    odds_ratio_lower = exp(estimate - 1.96 * standard_error),
    odds_ratio_upper = exp(estimate + 1.96 * standard_error)
  )

knitr::kable(
  parametric_table,
  digits = 3,
  caption = "Parametric coefficients and odds ratios from the final k = 50 GAM."
)

write_csv(
  parametric_table,
  "gam_month_parametric_effects.csv",
  na = "NA"
)
```

```{r final-model-diagnostics}
final_concurvity <- concurvity(
  gam_presence,
  full = TRUE
)

final_concurvity
```

```{r final-residual-acf, fig.width=8, fig.height=6}
acf(
  residuals(gam_presence, type = "deviance"),
  main = "ACF of deviance residuals: final k = 50 model"
)
```

```{r final-calibration, fig.width=11, fig.height=7}
calibration_data <- receiver_month_presence %>%
  mutate(
    predicted_probability = predict(
      gam_presence,
      type = "response"
    ),
    probability_bin = cut(
      predicted_probability,
      breaks = seq(0, 1, by = 0.1),
      include.lowest = TRUE
    )
  ) %>%
  group_by(month_f, probability_bin) %>%
  summarise(
    observed_rate = mean(present),
    predicted_rate = mean(predicted_probability),
    n = n(),
    .groups = "drop"
  )

calibration_plot <- ggplot(
  calibration_data,
  aes(x = predicted_rate, y = observed_rate, size = n)
) +
  geom_point(alpha = 0.75) +
  geom_abline(
    intercept = 0,
    slope = 1,
    linetype = "dashed",
    colour = "red"
  ) +
  facet_wrap(
    ~ month_f,
    labeller = as_labeller(month_labels)
  ) +
  coord_equal(xlim = c(0, 1), ylim = c(0, 1)) +
  labs(
    x = "Mean predicted probability",
    y = "Observed presence rate",
    size = "Station-months"
  ) +
  theme_bw(base_size = 12)

calibration_plot
```

# Generate spatial predictions

Predictions are generated over a 1-km grid within the convex hull of the
analysis receivers and are masked to remove land. Predictions represent the
estimated probability that at least one tagged kelt was detected at a receiver
location during a year-month, conditional on the fitted monthly pattern.

```{r prediction-grid}
# Use one coordinate for each receiver station represented in the model
model_stations <- receiver_month_presence %>%
  distinct(
    station,
    lon,
    lat,
    x,
    y
  )

model_stations_sf <- model_stations %>%
  st_as_sf(
    coords = c("x", "y"),
    crs = crs_projected,
    remove = FALSE
  )

receiver_hull <- model_stations_sf %>%
  st_union() %>%
  st_convex_hull()

receiver_bbox <- st_bbox(receiver_hull)

grid_coordinates <- expand_grid(
  x = seq(
    receiver_bbox[["xmin"]],
    receiver_bbox[["xmax"]],
    by = prediction_resolution_m
  ),
  y = seq(
    receiver_bbox[["ymin"]],
    receiver_bbox[["ymax"]],
    by = prediction_resolution_m
  )
)

prediction_grid_sf <- grid_coordinates %>%
  st_as_sf(
    coords = c("x", "y"),
    crs = crs_projected,
    remove = FALSE
  )

inside_receiver_hull <- lengths(
  st_within(prediction_grid_sf, receiver_hull)
) > 0

prediction_grid_sf <- prediction_grid_sf[inside_receiver_hull, ]

# Natural Earth land is used only to mask prediction cells over land.
land <- rnaturalearth::ne_countries(
  scale = "large",
  returnclass = "sf"
) %>%
  filter(admin %in% c("Canada", "United States of America")) %>%
  st_transform(crs_projected) %>%
  st_make_valid()

land_crop <- suppressWarnings(
  st_crop(land, receiver_bbox)
)

on_land <- lengths(
  st_intersects(prediction_grid_sf, land_crop)
) > 0

prediction_grid_sf <- prediction_grid_sf[!on_land, ]

prediction_data <- expand_grid(
  grid_row = seq_len(nrow(prediction_grid_sf)),
  month = months_keep
) %>%
  left_join(
    prediction_grid_sf %>%
      st_drop_geometry() %>%
      mutate(grid_row = row_number()),
    by = "grid_row"
  ) %>%
  mutate(
    month_f = factor(month, levels = months_keep)
  )

prediction_data$predicted_probability <- predict(
  gam_presence,
  newdata = prediction_data,
  type = "response"
)

stopifnot(!anyNA(prediction_data$predicted_probability))
```

# Plot monthly predicted habitat use

```{r prediction-map, fig.width=14, fig.height=10}
habitat_areas_plot <- habitat_areas %>%
  st_transform(crs_projected) %>%
  st_make_valid()

prediction_map <- ggplot() +
  geom_raster(
    data = prediction_data,
    aes(
      x = x,
      y = y,
      fill = predicted_probability
    )
  ) +
  geom_sf(
    data = land_crop,
    fill = "grey80",
    colour = "grey45",
    linewidth = 0.2
  ) +
  geom_sf(
    data = habitat_areas_plot,
    fill = NA,
    colour = "black",
    linewidth = 0.6
  ) +
  facet_wrap(
    ~ month_f,
    ncol = 4,
    labeller = as_labeller(month_labels)
  ) +
  coord_sf(
    crs = crs_projected,
    xlim = c(receiver_bbox[["xmin"]], receiver_bbox[["xmax"]]),
    ylim = c(receiver_bbox[["ymin"]], receiver_bbox[["ymax"]]),
    expand = FALSE
  ) +
  scale_fill_gradientn(
    colours = c(
      "#00FFFF",
      "lightgreen",
      "#FFFF00",
      "#FFA500",
      "#FF00A0"
    ),
    limits = c(0, 1),
    oob = squish,
    labels = label_number(accuracy = 0.1),
    name = "Predicted presence\nP(detection ≥ 1)"
  ) +
  labs(
    x = "Easting (m)",
    y = "Northing (m)"
  ) +
  theme_bw(base_size = 12) +
  theme(
    panel.grid = element_blank(),
    panel.background = element_rect(fill = "transparent", colour = NA),
    plot.background = element_rect(fill = "transparent", colour = NA),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    legend.position = "bottom",
    legend.key.width = grid::unit(3, "cm"),
    legend.text = element_text(size = 15),
    legend.title = element_text(size = 14),
    legend.background = element_rect(fill = "transparent", colour = NA),
    strip.background = element_blank(),
    strip.text = element_text(face = "bold", size = 15)
  )

prediction_map

ggsave(
  "gam_monthly_predicted_habitat_use.png",
  prediction_map,
  width = 14,
  height = 10,
  units = "in",
  dpi = 600,
  bg = "white"
)

ggsave(
  "gam_monthly_predicted_habitat_use.svg",
  prediction_map,
  width = 14,
  height = 10,
  units = "in",
  bg = "white"
)
```

# Compare predicted habitat use with raw detections

Raw detections are summarized as the number of unique tagged kelts detected at
each station during each retained calendar month.

```{r raw-detection-map, fig.width=15, fig.height=7.5, fig.cap="Monthly distribution of raw inner Bay of Fundy Atlantic salmon kelt detections. Open circles show locations where tagged kelts were detected, with symbol size corresponding to the number of unique fish detected. Black polygons delineate DFO marine habitat areas 3–8."}
# Count unique fish detected at each station by calendar month
raw_detection_counts <- det_analysis %>%
  filter(
    month %in% months_keep
  ) %>%
  group_by(
    station,
    month
  ) %>%
  summarise(
    n_unique_fish = n_distinct(fish),
    .groups = "drop"
  ) %>%
  left_join(
    receivers,
    by = "station"
  ) %>%
  filter(
    !is.na(lon),
    !is.na(lat)
  ) %>%
  mutate(
    month_f = factor(
      month.name[month],
      levels = month.name[months_keep]
    )
  )

# Convert detection locations to a spatial object
raw_detection_sf <- raw_detection_counts %>%
  st_as_sf(
    coords = c("lon", "lat"),
    crs = crs_geographic,
    remove = FALSE
  )

# Prepare land and habitat polygons in geographic coordinates
land_raw_plot <- land_crop %>%
  st_transform(crs_geographic)

habitat_areas_raw_plot <- habitat_areas %>%
  st_transform(crs_geographic) %>%
  st_make_valid()

# Define a common map extent from the primary receiver grid
longitude_limits <- range(
  receivers$lon,
  na.rm = TRUE
) + c(-0.08, 0.08)

latitude_limits <- range(
  receivers$lat,
  na.rm = TRUE
) + c(-0.05, 0.05)

# Select legend breaks supported by the data
maximum_fish <- max(
  raw_detection_counts$n_unique_fish,
  na.rm = TRUE
)

size_breaks <- c(1, 3, 5, 7, 9)

size_breaks <- size_breaks[
  size_breaks <= maximum_fish
]

if (!maximum_fish %in% size_breaks) {
  size_breaks <- sort(
    unique(c(size_breaks, maximum_fish))
  )
}

# Plot monthly raw detections
raw_detection_map <- ggplot() +
  geom_sf(
    data = land_raw_plot,
    fill = "grey88",
    colour = "grey45",
    linewidth = 0.25
  ) +
  geom_sf(
    data = habitat_areas_raw_plot,
    fill = NA,
    colour = "black",
    linewidth = 0.65
  ) +
  geom_sf(
    data = raw_detection_sf,
    aes(size = n_unique_fish),
    shape = 21,
    fill = "white",
    colour = "black",
    stroke = 0.8,
    alpha = 0.95
  ) +
  facet_wrap(
    ~ month_f,
    ncol = 4
  ) +
  coord_sf(
    crs = crs_geographic,
    xlim = longitude_limits,
    ylim = latitude_limits,
    expand = FALSE
  ) +
  scale_size_continuous(
    name = "Number of unique\nfish detected",
    range = c(2, 8),
    breaks = size_breaks,
    limits = c(1, maximum_fish)
  ) +
  scale_x_continuous(
    breaks = pretty_breaks(n = 4),
    labels = label_number(
      accuracy = 0.1,
      suffix = "°W",
      scale = -1
    )
  ) +
  scale_y_continuous(
    breaks = pretty_breaks(n = 4),
    labels = label_number(
      accuracy = 0.1,
      suffix = "°N"
    )
  ) +
  labs(
    x = "Longitude",
    y = "Latitude"
  ) +
  theme_bw(base_size = 14) +
  theme(
    panel.grid = element_blank(),

    # Remove the black border around each facet
    panel.border = element_blank(),

    panel.background = element_rect(
      fill = "white",
      colour = NA
    ),
    strip.background = element_blank(),
    strip.text = element_text(
      size = 16,
      face = "bold"
    ),
    axis.title = element_text(size = 15),
    axis.text = element_text(size = 12),
    axis.ticks = element_line(colour = "black"),
    legend.position = "right",
    legend.title = element_text(
      size = 14,
      face = "bold",
      hjust = 0.5
    ),
    legend.text = element_text(size = 12),
    legend.key.height = grid::unit(0.8, "cm"),
    legend.key.width = grid::unit(0.8, "cm")
  ) +
  guides(
    size = guide_legend(
      title.position = "top",
      title.hjust = 0.5,
      override.aes = list(
        shape = 21,
        fill = "white",
        colour = "black",
        alpha = 1
      )
    )
  )

raw_detection_map

ggsave(
  filename = "gam_monthly_raw_detections.png",
  plot = raw_detection_map,
  width = 15,
  height = 7.5,
  units = "in",
  dpi = 600,
  bg = "white"
)

ggsave(
  filename = "gam_monthly_raw_detections.svg",
  plot = raw_detection_map,
  width = 15,
  height = 7.5,
  units = "in",
  bg = "white"
)
```

# Export analysis products

```{r exports}
write_csv(
  receiver_month_presence,
  "gam_receiver_month_presence.csv",
  na = "NA"
)

write_csv(
  prediction_data,
  "gam_monthly_spatial_predictions.csv",
  na = "NA"
)

write_csv(
  prediction_comparison,
  "gam_basis_prediction_comparison.csv",
  na = "NA"
)

write_csv(
  model_comparison,
  "gam_basis_model_comparison.csv",
  na = "NA"
)

saveRDS(
  gam_presence,
  "gam_presence_model.rds"
)

capture.output(
  sessionInfo(),
  file = "gam_session_info.txt"
)
```
