---
title: "Temporal predation patterns"
author: "Caliyena Brown"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center",
  dpi = 300
)

library(dplyr)
library(tidyr)
library(readr)
library(lubridate)
library(ggplot2)
library(patchwork)
library(scales)
```

# Purpose

This document identifies the timing of inferred predation events, quantifies
the interval between the last confirmed alive detection and the first
post-predation detection, and summarizes inferred predation events by month.

# Import and prepare data

```{r import-data}
det <- readRDS("det_clean_no_release_site_det.rds")
meta_tag <- readRDS("meta_clean.rds")
predation_events <- readRDS("predation_events_clean.rds")

# Convert tag-level metadata to one biological record per fish.
meta <- meta_tag %>%
  select(fish, sex, tl, fl, release_dt) %>%
  distinct() %>%
  mutate(
    fish = as.integer(fish),
    release_dt = as.POSIXct(release_dt, tz = "UTC"),
    cohort = case_when(
      month(release_dt) == 4 ~ "April",
      month(release_dt) == 12 ~ "December",
      TRUE ~ NA_character_
    ),
    cohort = factor(cohort, levels = c("April", "December"))
  )

stopifnot(
  nrow(meta) == 101,
  anyDuplicated(meta$fish) == 0
)

# Confirm the required detection columns are present.
stopifnot(
  all(c("fish", "temp", "dt") %in% names(det))
)

# Parse detection times as UTC, then convert them to Halifax local time without
# changing the underlying instant represented by each timestamp.
if (inherits(det$dt, "POSIXct")) {
  det$dt <- with_tz(det$dt, tzone = "America/Halifax")
} else {
  det$dt <- with_tz(
    ymd_hms(det$dt, tz = "UTC", quiet = TRUE),
    tzone = "America/Halifax"
  )
}

det <- det %>%
  mutate(
    fish = as.integer(fish),
    temp = as.numeric(temp)
  ) %>%
  filter(
    !is.na(fish),
    !is.na(dt),
    fish %in% meta$fish
  ) %>%
  arrange(fish, dt)
```

# Prepare inferred predation events

The predation events used here were identified and validated in the
predation-mortality analysis. The preceding valid temperature record was used
to describe the temperature increase, and the immediately preceding detection
from any transmitter belonging to the same fish was used to quantify temporal
uncertainty.

```{r prepare-events}
# Convert predation timestamps to Halifax local time, matching the detections.
if (inherits(predation_events$predation_dt, "POSIXct")) {
  predation_events$predation_dt <- with_tz(
    predation_events$predation_dt,
    tzone = "America/Halifax"
  )
} else {
  predation_events$predation_dt <- with_tz(
    ymd_hms(predation_events$predation_dt, tz = "UTC", quiet = TRUE),
    tzone = "America/Halifax"
  )
}

predation_events <- predation_events %>%
  mutate(fish = as.integer(fish))

# Identify the preceding valid temperature record for every detection.
temperature_detections <- det %>%
  filter(!is.na(temp)) %>%
  group_by(fish) %>%
  arrange(dt, .by_group = TRUE) %>%
  mutate(
    previous_temperature_dt = lag(dt),
    previous_temp = lag(temp),
    temperature_increase = temp - previous_temp
  ) %>%
  ungroup()

# Add the preceding temperature record to each finalized predation event.
predation_events <- predation_events %>%
  select(fish, predation_dt, predation_temp) %>%
  left_join(
    temperature_detections %>%
      select(
        fish,
        detection_dt = dt,
        previous_temperature_dt,
        previous_temp,
        temperature_increase
      ),
    by = c("fish", "predation_dt" = "detection_dt")
  )

# Identify the immediately preceding detection of any transmitter belonging to
# the fish. This timestamp defines the temporal uncertainty interval.
last_alive <- det %>%
  inner_join(
    predation_events %>% select(fish, predation_dt),
    by = "fish"
  ) %>%
  filter(dt < predation_dt) %>%
  group_by(fish) %>%
  slice_max(dt, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  select(
    fish,
    last_alive_dt = dt
  )

# Add uncertainty intervals, release cohort, and predation month.
predation_events <- predation_events %>%
  left_join(last_alive, by = "fish") %>%
  left_join(
    meta %>% select(fish, cohort),
    by = "fish"
  ) %>%
  mutate(
    uncertainty_hours = as.numeric(
      difftime(predation_dt, last_alive_dt, units = "hours")
    ),
    uncertainty_days = uncertainty_hours / 24,
    predation_month = month(
      predation_dt,
      label = TRUE,
      abbr = FALSE
    )
  ) %>%
  arrange(predation_dt)

stopifnot(
  nrow(predation_events) == 13,
  anyDuplicated(predation_events$fish) == 0,
  all(predation_events$predation_temp > 20),
  all(!is.na(predation_events$last_alive_dt))
)

predation_events
```

# Temporal uncertainty

```{r temporal-uncertainty}
temporal_uncertainty_summary <- predation_events %>%
  summarise(
    n_events = n(),
    median_hours = median(uncertainty_hours, na.rm = TRUE),
    minimum_hours = min(uncertainty_hours, na.rm = TRUE),
    maximum_hours = max(uncertainty_hours, na.rm = TRUE),
    mean_hours = mean(uncertainty_hours, na.rm = TRUE)
  )

temporal_event_table <- predation_events %>%
  transmute(
    Fish = fish,
    `Release cohort` = as.character(cohort),
    `Last confirmed alive detection` = last_alive_dt,
    `Previous temperature detection` = previous_temperature_dt,
    `Preceding temperature (°C)` = round(previous_temp, 1),
    `Approximate predation detection` = predation_dt,
    `Predation temperature (°C)` = round(predation_temp, 1),
    `Temperature increase (°C)` = round(temperature_increase, 1),
    `Uncertainty interval (hours)` = round(uncertainty_hours, 1),
    `Uncertainty interval (days)` = round(uncertainty_days, 1),
    Month = as.character(predation_month)
  )

knitr::kable(
  temporal_uncertainty_summary,
  digits = 1,
  caption = paste(
    "Summary of the intervals between the last confirmed alive detection",
    "and first detection consistent with inferred predation."
  )
)

knitr::kable(
  temporal_event_table,
  align = "c",
  caption = paste(
    "Timing and temporal uncertainty of inferred predation events in",
    "Atlantic salmon kelts."
  )
)

write_csv(
  temporal_event_table,
  "temporal_predation_event_summary.csv",
  na = "NA"
)
```

# Temperature histories

```{r temperature-histories, fig.width=10, fig.height=13, fig.cap="Temperature histories of Atlantic salmon kelts with inferred predation events. Each panel represents one fish. Black crosses identify the first detection above 20°C for which the temperature history was considered consistent with inferred predation."}
predated_fish <- predation_events$fish

temperature_plot_data <- temperature_detections %>%
  filter(fish %in% predated_fish) %>%
  mutate(
    fish = factor(
      fish,
      levels = sort(unique(fish))
    )
  ) %>%
  arrange(temp)

predation_plot_data <- predation_events %>%
  mutate(
    fish = factor(
      fish,
      levels = levels(temperature_plot_data$fish)
    )
  )

temperature_history_plot <- ggplot(
  temperature_plot_data,
  aes(x = dt, y = temp, colour = temp, group = fish)
) +
  geom_line(
    linewidth = 0.5,
    alpha = 0.6,
    colour = "grey55"
  ) +
  geom_point(size = 2.2, alpha = 0.75) +
  geom_point(
    data = predation_plot_data,
    aes(x = predation_dt, y = predation_temp),
    inherit.aes = FALSE,
    shape = 4,
    colour = "black",
    size = 4.5,
    stroke = 1.1
  ) +
  facet_wrap(~ fish, scales = "free_x", ncol = 3) +
  scale_colour_viridis_c(
    option = "plasma",
    limits = c(0, 40),
    oob = scales::squish,
    name = "Temperature (°C)",
    guide = guide_colourbar(
      barheight = grid::unit(4.5, "cm")
    )
  ) +
  scale_y_continuous(
    limits = c(0, 40),
    breaks = seq(0, 40, by = 10)
  ) +
  scale_x_datetime(date_labels = "%d-%b") +
  labs(
    x = "Date (day–month)",
    y = "Temperature (°C)"
  ) +
  theme_bw(base_size = 12) +
  theme(
    strip.background = element_blank(),
    strip.text = element_text(size = 12, face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1),
    axis.title = element_text(size = 14),
    legend.title = element_text(size = 13),
    legend.text = element_text(size = 11)
  )

temperature_history_plot

ggsave(
  "predation_temperature_histories.png",
  temperature_history_plot,
  width = 10,
  height = 13,
  units = "in",
  dpi = 600,
  bg = "white"
)
```

# Monthly predation patterns

Predation events are summarized by month of year. Fish availability represents
the number of detected fish under observation during at least part of each
month. Predated fish remain available through the month containing their
inferred predation event; other detected fish remain available through the
month of their final detection.

```{r monthly-data}
month_levels <- month.name

# Create a complete 12-month reference table
month_grid <- tibble(
  month = month_levels
)

# Count inferred predation events by month
predation_month_counts <- predation_events %>%
  mutate(
    month = as.character(predation_month)
  ) %>%
  count(
    month,
    name = "n_predation"
  )

# Establish the final observation time for each detected fish
last_detections <- det %>%
  group_by(fish) %>%
  summarise(
    last_detection_dt = max(dt),
    .groups = "drop"
  )

# Determine the months during which each detected fish was available
availability_data <- meta %>%
  inner_join(
    last_detections,
    by = "fish"
  ) %>%
  left_join(
    predation_events %>%
      select(fish, predation_dt),
    by = "fish"
  ) %>%
  mutate(
    release_dt_local = with_tz(
      release_dt,
      tzone = "America/Halifax"
    ),

    # Predated fish end at predation; other fish end at last detection
    endpoint_dt = if_else(
      !is.na(predation_dt),
      predation_dt,
      last_detection_dt
    ),

    start_month = as.Date(
      floor_date(release_dt_local, unit = "month")
    ),

    end_month = as.Date(
      floor_date(endpoint_dt, unit = "month")
    )
  ) %>%
  filter(end_month >= start_month) %>%
  rowwise() %>%
  mutate(
    month_sequence = list(
      seq.Date(
        from = start_month,
        to = end_month,
        by = "month"
      )
    )
  ) %>%
  ungroup() %>%
  unnest(month_sequence) %>%
  mutate(
    # Store month as character to avoid ordered-factor conflicts
    month = month.name[lubridate::month(month_sequence)]
  )

# Count fish available during each month
available_month_counts <- availability_data %>%
  count(
    month,
    name = "n_available"
  )

# Combine counts with the complete 12-month reference table
monthly_summary <- month_grid %>%
  left_join(
    predation_month_counts,
    by = "month"
  ) %>%
  left_join(
    available_month_counts,
    by = "month"
  ) %>%
  mutate(
    n_predation = replace_na(n_predation, 0L),
    n_available = replace_na(n_available, 0L),

    events_per_100_available = if_else(
      n_available > 0,
      100 * n_predation / n_available,
      NA_real_
    ),

    # Convert month to an ordered plotting factor only after all joins
    month = factor(
      month,
      levels = month_levels,
      ordered = TRUE
    )
  )

knitr::kable(
  monthly_summary,
  digits = 1,
  caption = paste(
    "Monthly inferred predation events and numbers of detected kelts",
    "available for observation."
  )
)

write_csv(
  monthly_summary,
  "monthly_predation_summary.csv",
  na = "NA"
)
```

```{r monthly-figure, fig.width=10, fig.height=6.5, fig.cap="Monthly timing of inferred predation events and numbers of detected Atlantic salmon kelts available for observation. Bars show the number of inferred predation events, and the black line shows the number of fish available during each month."}

# Colours used for monthly predation-event bars
month_colours <- c(
  "January" = "#D9D9D9",
  "February" = "#D9D9D9",
  "March" = "#D9D9D9",
  "April" = "#E4A3A9",
  "May" = "#D3D3D3",
  "June" = "#D8576E",
  "July" = "#FFC000",
  "August" = "#4EA72E",
  "September" = "#13501B",
  "October" = "#D9D9D9",
  "November" = "#D9D9D9",
  "December" = "#D9D9D9"
)

# Calculate the scaling needed to display fish availability on the same plot
scale_factor <- max(
  monthly_summary$n_available,
  na.rm = TRUE
) / max(
  monthly_summary$n_predation,
  na.rm = TRUE
)

# Scale fish availability to the range of the left axis
monthly_plot_data <- monthly_summary %>%
  mutate(
    n_available_scaled = n_available / scale_factor
  )

# Create the combined monthly plot
monthly_timing_plot <- ggplot(
  monthly_plot_data,
  aes(x = month)
) +
  # Predation-event bars
  geom_col(
    aes(
      y = n_predation,
      fill = month
    ),
    width = 0.75,
    colour = "black",
    linewidth = 0.3
  ) +
  # Fish-availability line
  geom_line(
    aes(
      y = n_available_scaled,
      group = 1
    ),
    linewidth = 1.2,
    colour = "black"
  ) +
  geom_point(
    aes(y = n_available_scaled),
    size = 3,
    shape = 21,
    fill = "black",
    colour = "black",
    stroke = 1
  ) +
  scale_fill_manual(
    values = month_colours,
    guide = "none"
  ) +
  scale_y_continuous(
    name = "Number of inferred predation events",
    breaks = scales::breaks_width(1),
    expand = expansion(mult = c(0, 0.08)),

    # Convert the scaled line back to fish counts on the right axis
    sec.axis = sec_axis(
      trans = ~ . * scale_factor,
      name = "Number of fish available",
      breaks = scales::breaks_pretty()
    )
  ) +
  labs(
    x = "Month"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    axis.text.x = element_text(
      angle = 45,
      hjust = 1
    ),
    axis.title.x = element_text(
      size = 16,
      margin = margin(t = 10)
    ),
    axis.title.y.left = element_text(
      size = 15,
      margin = margin(r = 10)
    ),
    axis.title.y.right = element_text(
      size = 15,
      margin = margin(l = 10)
    ),
    axis.text = element_text(size = 13),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank()
  )

monthly_timing_plot

ggsave(
  filename = "monthly_predation_patterns.png",
  plot = monthly_timing_plot,
  width = 10,
  height = 6.5,
  units = "in",
  dpi = 600,
  bg = "white"
)
```

# Reproducibility

```{r session-info}
sessionInfo()
```
